The Design and Implementation of the Synchronous Language \CEU

### 2016-10-01 paper.tex
(>) soft real-time system

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with a simpler semantics and 
more fine-grained control over program execution.
%
\CEU uses an event-triggered notion of time that enables compile-time checks to 
detect conflicting concurrent statements, resulting in deterministic and 
concurrency-safe programs.
%
Using Esterel as a base, we present the particularities of our design, such as 
stack-based internal events, concurrency detection, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

Despite its success and influence, Esterel has a complex semantics that
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems%
%These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
A complex semantics not only challenges the analysis and compilation of 
programs, but also affects the programmer's understanding about the code, who,
ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of non-deterministic
execution for intra-reaction statements.
Non determinism prevents threads from interacting with stateful system calls safely 
and makes shared-memory concurrency not as straightforward as reading and 
writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU proposes a semantics with fine-grained control for intra-reaction
execution, and is amenable to concurrency checks that improve safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency checks to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.checks}).
%
\item Safe integration with \C that enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of the proposed semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

@@

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction of a given event (we discuss implementation
considerations in Sections~\ref{sec.impl.single}~and~\ref{sec.impl.environment}).
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulate for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, requiring \emph{worst case reaction time} 
analysis to infer appropriate tick lengths~\cite{esterel.fixed}.
% TODO-no: Such analysis is orthogonal

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
We consider that this approach is more familiar to programmers and simplifies 
the reasoning about concurrency.
%
As far as we know, \CEU is the first to encrust this event-triggered notion of 
time in the core of the language, which is a prerequisite for the concurrency 
checks that enable safe shared-memory concurrency.

In \CEU internal events support stack-based micro reactions within external
reactions, providing more fine-grained control for intra-reaction execution.
%
Some variants of Statecharts~\cite{statecharts} also distinguish internal from
external events~\cite{statecharts.variants}.
%Some variants of Statecharts support local events which imply handling intra
%reactions like Céu.
In Statemate, ``reactions to external and internal events (...) can be sensed only
after completion of the step''~\cite{statecharts.statemate}, implying queue-based
execution.
In Stateflow, ``the receiving state (of the event) acts here as a
function''~\cite{statecharts.stateflow}, which is similar to Céu's stack-based
execution.
To avoid recursion, Céu adopts delayed awaits, while in Statemate, ``loops in
the broadcasting of events (between states) are forbidden''.

@@

We present the design and implementation of \CEU, a synchronous reactive 
language inspired by Esterel with simpler semantics and more fine-grained 
control for intra-reaction execution.

>>>>

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with a simpler semantics and 
more fine-grained control over program execution targeting soft real-time
systems.
%
\CEU uses an event-triggered notion of time that enables compile-time checks to 
detect conflicting concurrent statements, resulting in deterministic and 
concurrency-safe programs.
%
Using Esterel as a base, we present the particularities of our design, such as 
stack-based internal events, concurrency detection, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

Despite its success and influence, Esterel has a complex semantics that
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems%
%These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
A complex semantics not only challenges the analysis and compilation of 
programs, but also affects the programmer's understanding about the code, who,
ultimately, has to solve the errors when facing corner cases.
%
Furthermore, Esterel's semantics is non-deterministic for intra-reaction
statements, which prevents threads from interacting with stateful system calls
safely and makes shared-memory concurrency not as straightforward as reading
and writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU proposes a semantics for soft real-time systems with fine-grained control
for intra-reaction execution, which is amenable to concurrency checks that
improve safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency checks to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.checks}).
%
\item Safe integration with \C that enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of the proposed semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

The semantics of \CEU implies an implementation with a runtime stack and a
sequential scheduler.
In contrast with Esterel's semantics, this prevents programs to be compiled
directly and efficiently to hardware~\cite{esterel.constructive}.
It also makes difficult to statically determine worst-case reaction times
for hard real-time systems~\cite{esterel.fixed}.

@@

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction of a given event (we discuss implementation
considerations in Sections~\ref{sec.impl.single}~and~\ref{sec.impl.environment}).
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulate for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, requiring \emph{worst case reaction time} 
analysis to infer appropriate tick lengths~\cite{esterel.fixed}.
% TODO-no: Such analysis is orthogonal
%
For soft real-time systems, accumulating delays and occasionally postponing
reactions might be enough.
However, hard real-time systems require a more robust approach such as
determining fixed-length ticks that can be statically verified.

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
As far as we know, \CEU is the first to encrust this event-triggered notion of 
time in the core of the language, which is a prerequisite for the concurrency 
checks that enable safe shared-memory concurrency.

In \CEU, internal events support stack-based micro reactions within external
reactions, providing more fine-grained control for intra-reaction execution.
%
Some variants of Statecharts~\cite{statecharts} also distinguish internal from
external events~\cite{statecharts.variants}.
%Some variants of Statecharts support local events which imply handling intra
%reactions like Céu.
In Statemate, ``reactions to external and internal events (...) can be sensed only
after completion of the step''~\cite{statecharts.statemate}, implying queue-based
execution.
In Stateflow, ``the receiving state (of the event) acts here as a
function''~\cite{statecharts.stateflow}, which is similar to Céu's stack-based
execution.
To avoid recursion, Céu adopts delayed awaits, while in Statemate, ``loops in
the broadcasting of events (between states) are forbidden''.

@@

We present the design and implementation of \CEU, a synchronous reactive 
language inspired by Esterel with event-driven semantics and more fine-grained 
control for intra-reaction execution.
%
On the one hand, this approach is familiar to programmers in general,
abstracting tick sampling with reactions to unique events from their
application domain.
On the other hand, this level of abstraction does not suit systems with
hard-real time requirements in which interacting directly with a concrete
notion of tick is more robust.
